---
title: "Assignment 2"
author: "Damian Ke and Kyriakos Papadopoulos"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 1. KERNEL METHODS
Implement a kernel method to predict the hourly temperatures for a date and place in Sweden.
To do so, you are provided with the files stations.csv and temps50k.csv. These
files contain information about weather stations and temperature measurements in the stations
at different days and times. The data have been kindly provided by the Swedish Meteorological
and Hydrological Institute (SMHI).
You are asked to provide a temperature forecast for a date and place in Sweden. The
forecast should consist of the predicted temperatures from 4 am to 24 pm in an interval of 2
hours. Use a kernel that is the sum of three Gaussian kernels:
Y The first to account for the physical distance from a station to the point of interest. For
this purpose, use the function distHaversine from the R package geosphere.
Y The second to account for the distance between the day a temperature measurement
was made and the day of interest.
Y The third to account for the distance between the hour of the day a temperature measurement
was made and the hour of interest.
Choose an appropriate smoothing coefficient or width for each of the three kernels above.
No cross-validation should be used. Instead, choose manually a width that gives large kernel
values to closer points and small values to distant points. Show this with a plot of the kernel
value as a function of distance. Help: Note that the file temps50k.csv may contain temperature
measurements that are posterior to the day and hour of your forecast. You must filter
such measurements out, i.e. they cannot be used to compute the forecast.
Finally, repeat the exercise above by combining the three kernels into one by multiplying
them, instead of summing them up. Compare the results obtained in both cases and elaborate
on why they may differ.
The only R package that is allowed to solve this assignment is the geosphere package
(specifically, the function distHaversine). Feel free to use the template below to solve the
assignment.

```{r, echo=FALSE}
set.seed(1234567890)
library(ggplot2)
library(geosphere)
#Else could not read the file, required fileEncoding
#Changed to read.csv2 to be able to read.
stations <- read.csv2("stations.csv",fileEncoding = "windows-1258")
temps <- read.csv("temps50k.csv")
st <- merge(stations,temps,by="station_number")


# These three values are up to the students
h_distance <- 20
h_date <- 30
h_time <- 2

#The point to predict (up to the students) - Falun-Lugnet
a <- 60.619 
b <- 15.6603

date_predicted <- "2007-10-08" # The date to predict (up to the students)
times <- c("04:00:00", "06:00:00", "08:00:00","10:00:00","12:00:00","14:00:00",
           "16:00:00","18:00:00","20:00:00","22:00:00","24:00:00")

# Students’ code here
#Filtering data
filtered_data = subset(st, as.Date(date) < as.Date(date_predicted))

#Converting columns to correct type
filtered_data$time = strptime(filtered_data$time, format="%H:%M:%S")
filtered_data$longitude = as.numeric(filtered_data$longitude)
filtered_data$latitude = as.numeric(filtered_data$latitude)


#Adding 3 Kernels

#Kernel 1
#The first to account for the physical distance from a station to the point of interest. For
#this purpose, use the function distHaversine from the R package geosphere.

#Calculated in meters --> Divided by 1000 to get in km
distance_difference = distHaversine(c(b,a),rev(filtered_data[,4:5]))/1000


#Kernel 2
# The second to account for the distance between the day a temperature measurement
#was made and the day of interest.
day_difference = as.numeric(as.Date(date_predicted)-as.Date(filtered_data$date))%%365

#Kernel 3
#The third to account for the distance between the hour of the day a temperature measurement was made 
#and the hour of interest.
time_difference = matrix(nrow=nrow(filtered_data),ncol=length(times))
i=1
#For loop to calculate the difference for each hourly temperature.
for (time in times){
  time_difference[,i] = abs(as.integer(difftime(strptime(time,format="%H:%M:%S"),
                                               filtered_data$time, units="hours")))
  
  i=i+1
}
#Issue that prediction goes up for time difference after 12.
time_difference = ifelse(time_difference >12, 24-time_difference, time_difference)

#Calculations of Gaussian kernels with kernel trick
results_distance = exp(((-distance_difference**2)/(2*(h_distance**2))))
results_day = exp(((-day_difference**2)/(2*(h_date**2))))
results_time = exp(((-time_difference**2)/(2*(h_time**2))))

#Addition Predictions
pred_added = (results_distance+results_day+results_time)

results_added= matrix()
#Loop to calculate the prediction for each hour.
for(i in 1:length(times)){
  results_added[i] = sum(pred_added[,i]*filtered_data$air_temperature)/sum(pred_added[,i])
}


#Multiplied
pred_multiplied = (results_distance*results_day*results_time)

results_multiplied = matrix()
for(i in 1:length(times)){
  results_multiplied[i] = sum(pred_multiplied[,i]*filtered_data$air_temperature)/sum(pred_multiplied[,i])
}

#Plot the results
plot_df_sum_add = data.frame(results_added, results_multiplied,times)
colnames(plot_df_sum_add) = c("Added", "Multiplied","Times")

plot_df_sum_add = ggplot(plot_df_sum_add, aes(x=as.factor(substr(Times,1,5))))+
  geom_point(aes(y=Added, color="Added Kernel"))+
  geom_point(aes(y=Multiplied, color="Multiplied Kernel"))+
  xlab("Time")+
  ylab("Prediction")+
  ggtitle("Prediction for Added and Multiplied Kernels")+
  scale_color_manual(name="Definitions", values=c("Added Kernel"="red", 
                                                  "Multiplied Kernel"="blue"))

```

## Question 1
Choose manually a width that gives large kernel values to closer points and small values to distant points
Show this with a plot of the kernel value as a function of distance.


```{r, echo=FALSE}
#Plot of results vs distance
plot_df_3 = data.frame(results_distance, distance_difference)
colnames(plot_df_3) = c("Value", "Distance")

ggplot(plot_df_3, aes(x=Distance,y=Value))+
  geom_line()+
  xlab("Distance Difference")


plot_df_4 = data.frame(results_day, day_difference)
colnames(plot_df_4) = c("Value", "Distance")

ggplot(plot_df_4, aes(x=Distance,y=Value))+
  geom_line()+
  xlab("Day Difference")


plot_df_5 = data.frame(results_time[,11], time_difference[,11])
colnames(plot_df_5) = c("Value", "Distance")

ggplot(plot_df_5, aes(x=Distance,y=Value))+
  geom_line()+
  xlab("Time Difference")

```

**Answer**

The chosen kernels widths are:
Distance : `r h_distance`

Date : `r h_date`

Time : `r h_time`

Important to mention that distance was divided by 1000 to get the distance into Kilometers.
Therefore, the width of distance was set to 20km which seemed to give reasonable values.

The date was divided into modulo 365 to better predict the impact of each month.
The width was set to 30 so it can find a difference for each month which gave good prediction.
For this exercise, the date was also tested on total date difference. But it seemed
that it gave worse results than the earlier option, therefore modulo of 365 was used.

Lastly, for time difference if values were larger than 12, then it would calculate 24-time difference to get
correct values. Else there would be an increase of temperature at about hours 18-24.
The width of time was set to 2, which seemed to be better at predicting the temperature
at different times.

## Question 2 
Compare the results obtained in both cases and elaborate
on why they may differ.

```{r, echo=FALSE}
plot_df_sum_add
```

**Answer**
Chosen location is (60.619, 15.6603) which is Falun-Lugnet.
The predicted date is "2007-10-08". The chosen location and date was used to predict
the actual temperature within the dataset. 
For date "2007-10-08" at time 06.00 the temperature was 6.9 degrees.
Which shows that both methods were quite close to the predicted value.

As it can be seen in the figure, the multiplied kernel has higher prediction values than
the added one. Also there seem to be smaller temperature variation in the added kernel, meanwhile
the multiplied one has higher variance. 
For added kernel, each kernel has their own value which is not impacted by values from the other
kernels. For the multiplied kernel, each kernel impact the value of the others as they are multiplicated.
Therefore, values of the kernels has a bigger impact on the total prediction which is
why there is a difference between these kernels.

In addition, kernels width have impact on these results, as the graph lines in question 1 impact the total prediction.
For larger values of the width, additional kernel differences will have larger kernel values and will therefore 
impact the kernel.
Meanwhile for small values of width, the large kernel values will have less value and therefore
have smaller impact on the kernel.

