---
title: "Assignment 2"
author: "Damian Ke and Kyriakos Papadopoulos"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Assignment 2: Bootstrap, jackknife and confidence intervals

#Question 1
Plot the histogram of Price. Does it remind any conventional distribution? Compute the
mean price.
```{r}
library(ggplot2)



#Assignment 2:
df = read.csv2("prices1.csv")
p = ggplot(df, aes(x=Price)) +
  geom_histogram(binwidth=150)
p

mean_price = mean(df$Price)

#Looks like gamma?


```

**Answer**
The di

#Question 2
Estimate the distribution of the mean price of the house using bootstrap. Determine the
bootstrap bias–correction and the variance of the mean price. Compute a 95% confidence
interval for the mean price using bootstrap percentile, bootstrap BCa, and first–order
normal approximation.
```{r}
library(boot)

set.seed(12345)
stat1 = function(data,vn){
  mean(data[vn,]$Price)
}


res=boot(df,stat1,R=1000)
res


###Determine the bootstrap bias - correction and the variance of the mean price.
#Bootstrap bias corrections in slides and variance estimation in slides.
B=1000 #Q: As R --> 1000?


#Q:I got lost on Krzysztof slides the estimator w = T(D)
# Then what is Bar T(D*) -->
#Q:#Whats res$t0 exactly? As mean(res$t) is similar to res$t0.
#Should i use mean(res$t) or just res$t0?
t_1 = 2*res$t0-mean(res$t)

#Variance
var_t = (1/(B-1))*(sum((res$t-mean(res$t))^2))


#Q: Is this enough to show percentile, bca and normal? As basic is within?
print(boot.ci(res))

plot(res)

```

**Answer**

#Question 3
Estimate the variance of the mean price using the jackknife and compare it with the
bootstrap estimate

```{r}
n = length(df$Price)
#n = B
#Jackknife variance
#Q: According to Oleg - T(D_1) = Mean of the sample 1. 
#Does it also work for jackknife then or should we use other estimator?
#Length of the datasample
#Q: Should the n for jackknife variance be equal as 2022 lecture = B
#Or as in oleg that it is equal to length of the sample (Where we say that k =1)
#(So r = n)
t_j = rep(0,n)
for (j in 1:n){
  t_j[j] = n*mean(df$Price)-(n-1)*mean(df$Price[-j])
}
j_t = mean(t_j)

var_t3 = (1/(n*(n-1)))*sum((t_j-j_t)**2)


```

**Answer*


##Question 4
Compare the confidence intervals obtained with respect to their length and the location of
the estimated mean in these intervals.
```{r}
#4 - Look at it
interval = boot.ci(res)
interval_percentile = interval$percent[1,4:5]
interval_bca = interval$bca[1,4:5]
interval_normal = interval$normal[1,2:3]

#As it states confidence intervals obtained with respect to their length.
#Small Q: As Confidence intervals were calculated by inbuilt formula.
#Is this enough for this assignment?

p_percentile = ggplot(df, aes(x=Price)) +
  geom_histogram(binwidth=150)+
  geom_vline(xintercept=interval_percentile[1:2])+
  geom_vline(xintercept=t_1)
p_percentile

p_bca =  ggplot(df, aes(x=Price)) +
  geom_histogram(binwidth=150)+
  geom_vline(xintercept=interval_bca[1:2])+
  geom_vline(xintercept=t_1)
p_bca

p_normal =  ggplot(df, aes(x=Price)) +
  geom_histogram(binwidth=150)+
  geom_vline(xintercept=interval_normal[1:2])+
  geom_vline(xintercept=t_1)
p_normal


```

**Answer**


